ANGSD-wrapper: utilities for analyzing next generation sequencing data

### 2016-04-28 manuscript.tex
final^5
fixed typos

High throughput sequencing has revolutionized evolutionary genetics, allowing researchers to quickly assay large numbers of individuals and interrogate genome-wide variation. 
Application of these approaches has led to changes in both experimental design and data analysis \citep{ekblom2011applications}. 
Many popular software packages used by researchers for analysis of comparative resequencing data \citep[see][]{excoffier2006computer} were not designed to handle these novel data types or efficiently analyze the large volumes of data now being generated. 
Despite the decreasing cost of sequencing, researchers must nonetheless allocate finite resources and balance the depth of sequencing and the breadth of a sample. 
This poses a challenge for population genetics analysis, which generally requires accurate polymorphism calls in a broad sample \citep{pluzhnikov1996optimal, felsenstein2006accuracy}. 
While these experimental design challenges in molecular population genetic studies have existed for at least two decade, high throughput sequencing brings the added technical challenges of highly variable coverage, missing data, and high per-nucleotide error rates.

@@

Here we present ANGSD-wrapper, a user-friendly interface to ANGSD. 
ANGSD-wrapper takes the form of a set of configuration files and wrapper scripts (Figure \ref{fig:supp2}) that streamline the execution of multi-step pipelines required for data analysis in ANGSD. 
ANGSD-wrapper also eases the configuration of ANGSD-related programs such as ngsPopGen \citep{ngsPopGen}, ngsF\citep{vieira2013estimating}, and ngsAdmix \citep{pmid24026093}. 
As in ANGSD, ANGSD-wrapper allows users to perform whole genome analysis or analyze a set of user-defined windows across the genome.
The wrapper scripts are written against a frozen versions of ANGSD and supporting tools for consistency of analysis. 
Because the large volume of data associated with high throughput sequence analysis is often difficult to visualize, ANGSD-wrapper also provides a suite of interactive visualization tools to plot results and explore patterns at multiple scales.  
We demonstrate some of the analyses possible using ANGSD-wrapper when applied to low-coverage resequencing using data from domesticated maize and two related wild teosinte subspecies. 
ANGSD-wrapper is freely available from \url{https://github.com/mojaveazure/angsd-wrapper}.

@@

ANGSD-wrapper is divided into scripts associated with analytical approaches implemented in ANGSD and associated software. 
ANGSD-wrapper provides a common configuration file, \texttt{Common\_Config}, which holds variables that are likely to remain constant across analyses, including identifiers for chromosomal regions and the paths to project directories.
In ANGSD-wrapper, each method is self-contained in a shell script which uses information from the common configuration file and a method-specific configuration file. 
Each analysis is run using a simple command:

@@

\section*{Results and Discussion}
As a demonstration of analyses in ANGSD-wrapper, we explore patterns of pairwise nucleotide diversity in a single genomic region of domesticated maize and wild teosinte. 
We used a subset of the resequenced samples from the Maize HapMap2 project (Table \ref{tab:samples}) and calculated summary statistics using a 10Mb region on chromosome 10 \citep{chia2012maize}. 
The data are available at \url{https://figshare.com/articles/Example_Data_tar_bz2/2063442} and can be downloaded from within AGNSD-wrapper using teh command

@@

We first use the ``SFS'' method to estimate the site frequency spectrum (SFS) of both maize and its wild progenitor \textit{Zea mays} ssp. \textit{parviglumis}, assuming an inbreeding coefficient of $F=1$ for these highly inbred samples. 
The SFS and diversity statistics were calculated using ancestral states inferred in the ``Ancestral Sequence'' method from a single resequenced genome of \textit{Tripsacum dactyloides}.  
We show that the maize SFS is skewed towards more intermediate-frequency variants (Figure \ref{fig:figure3}A-B, Tajima's D of 0.2 and 0.0085 in maize and teosinte, respectively), likely a result of the bottleneck associated with maize domestication \citep{eyre1998investigation, Beissinger031666}.
Using the ``Thetas'' method we find further evidence of the domestication bottleneck, with mean levels of pairwise nucleotide diversity in this region in maize $\approx 25\%$ lower than in teosinte (0.0061 and 0.0082, respectively; Figure \ref{fig:figure3}C). 
Using the ``2DSFS method,'' which includes an \fst calculation, we find a mean \fst in this region of 0.116, nearly identical to the  genome-wide value of 0.11 reported in \cite{hufford2012comparative}.   
None of the genes in this region have been identified as potential domestication candidates \cite{hufford2012comparative}, consistent with the lack of extended regions of high \fst in our analysis (Figure \ref{fig:suppfst}).

\begin{figure}
\centering
\caption{Admixture analysis for {\it Zea mays} ssp. {\it mays}, {\it Zea mays} ssp. {\it parviglumis}, and {\it Zea mays} ssp. {\it mexicana} (\textit{mex}). Colors represent the proportion of each individual's genome assigned to one of the {\it K}=4 source populations. 
Individuals are shown in the same order as Table \ref{tab:samples}. For results using other values of \texit{K} see Figure \ref{fig:suppadmix}.}
\includegraphics[width=0.8\linewidth]{figures/mt4labeled.pdf}
\label{fig:admixture}
\end{figure}

>>>>

High throughput sequencing has revolutionized evolutionary genetics, allowing researchers to quickly assay large numbers of individuals and interrogate genome-wide variation. 
Application of these approaches has led to changes in both experimental design and data analysis \citep{ekblom2011applications}. 
Many popular software packages used by researchers for analysis of comparative resequencing data \citep[see][]{excoffier2006computer} were not designed to handle these novel data types or efficiently analyze the large volumes of data now being generated. 
Despite the decreasing cost of sequencing, researchers must nonetheless allocate finite resources and balance the depth of sequencing and the breadth of a sample. 
This poses a challenge for population genetics analysis, which generally requires accurate polymorphism calls in a broad sample \citep{pluzhnikov1996optimal, felsenstein2006accuracy}. 
While these experimental design challenges in molecular population genetic studies have existed for at least two decades, high throughput sequencing brings the added technical challenges of highly variable coverage, missing data, and high per-nucleotide error rates.

@@

Here we present ANGSD-wrapper, a user-friendly interface to ANGSD. 
ANGSD-wrapper takes the form of a set of configuration files and wrapper scripts (Figure \ref{fig:supp2}) that streamline the execution of multi-step pipelines required for data analysis in ANGSD. 
ANGSD-wrapper also eases the configuration of ANGSD-related programs such as ngsTools \citep{fumagalli2014ngstools}, ngsF\citep{vieira2013estimating}, ngsAdmix \citep{pmid24026093}, and PCA \citep{fumagalli2013quantifying}. 
As in ANGSD, ANGSD-wrapper allows users to perform whole genome analysis or analyze a set of user-defined windows across the genome.
The wrapper scripts are written against a frozen versions of ANGSD (v [INSERT VERSION]) and supporting tools for consistency of analysis. 
Because the large volume of data associated with high throughput sequence analysis is often difficult to visualize, ANGSD-wrapper also provides a suite of interactive visualization tools to plot results and explore patterns at multiple scales.  
We demonstrate some of the analyses possible using ANGSD-wrapper when applied to low-coverage resequencing using data from domesticated maize and two related wild teosinte subspecies. 
ANGSD-wrapper is freely available from \url{https://github.com/mojaveazure/angsd-wrapper}.

@@

ANGSD-wrapper is divided into scripts associated with analytical approaches implemented in ANGSD and associated software. 
It provides a common configuration file, \texttt{Common\_Config}, which holds variables that are likely to remain constant across analyses, including identifiers for chromosomal regions and the paths to project directories.
In ANGSD-wrapper, each method is self-contained in a shell script which uses information from the common configuration file and a method-specific configuration file. 
Each analysis is run using a simple command:

@@

\section*{Results and Discussion}
As a demonstration of analyses in ANGSD-wrapper, we explore patterns of pairwise nucleotide diversity in a single genomic region of domesticated maize and wild teosinte. 
We used a subset of the resequenced samples from the Maize HapMap2 project (Table \ref{tab:samples}) and calculated summary statistics using a 10Mb region on chromosome 10 \citep{chia2012maize}. 
The data are available at \url{https://figshare.com/articles/Example_Data_tar_bz2/2063442} and can be downloaded from within ANGSD-wrapper using the command

@@

We first use the ``SFS'' method to estimate the site frequency spectrum (SFS) of both maize and its wild progenitor \textit{Zea mays} ssp. \textit{parviglumis}, assuming an inbreeding coefficient of $F=1$ for these highly inbred samples. 
The SFS and diversity statistics were calculated using ancestral states inferred by the ``Ancestral Sequence'' method from a single resequenced genome of \textit{Tripsacum dactyloides}.  
We show that the maize SFS is skewed towards more intermediate-frequency variants (Figure \ref{fig:figure3}A-B, Tajima's D of 0.2 and 0.0085 in maize and teosinte, respectively), likely a result of the bottleneck associated with maize domestication \citep{eyre1998investigation, Beissinger031666}.
Using the ``Thetas'' method we find further evidence of the domestication bottleneck, with mean levels of pairwise nucleotide diversity in this region in maize $\approx 25\%$ lower than in teosinte (0.0061 and 0.0082, respectively; Figure \ref{fig:figure3}C). 
Using the ``2DSFS method,'' which includes an \fst calculation, we find a mean \fst in this region of 0.116, nearly identical to the  genome-wide value of 0.11 reported in \cite{hufford2012comparative}.   
None of the genes in this region have been identified as potential domestication candidates \cite{hufford2012comparative}, consistent with the lack of extended regions of high \fst in our analysis (Figure \ref{fig:suppfst}).

\begin{figure}
\centering
\caption{Admixture analysis for {\it Zea mays} ssp. {\it mays}, {\it Zea mays} ssp. {\it parviglumis}, and {\it Zea mays} ssp. {\it mexicana} (\textit{mex}). Colors represent the proportion of each individual's genome assigned to one of the {\it K}=4 source populations. 
Individuals are shown in the same order as Table \ref{tab:samples}. For results using other values of \textit{K} see Figure \ref{fig:suppadmix}.}
\includegraphics[width=0.8\linewidth]{figures/mt4labeled.pdf}
\label{fig:admixture}
\end{figure}
